Hippocampal segmentation with MAGeT

### 2013-02-06 paper.Rnw
Updated methods with ADNI1 description and minctracc parameters

\begin{figure}[h]
  \centering
    \includegraphics[width=0.5\textwidth]{figure/MAGeT-figure.png}
  \caption{Diagram of the MAGeT Brain algorithm}
\end{figure}

@@

\subsubsection{ADNI-1 1.5T Screening}
\todo{include image characteristics}
\todo{describe SNT manual segmentations}

<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",  caption="ADNI-1 1.5T Screening demographics")
@

@@

\subsubsection{SZ First Episode Patients}
\todo{include image characteristics}

<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics")
@

@@

\subsubsection{Automatic Normalization and Image Matching and Anatomical Labeling (ANIMAL)}
The ANIMAL algorithm carries out Image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the
whole brain \citep{Collins}. In the second phase, nonlinear registration is completed using
the ANIMAL algorithm \citep{Collins1995}: an iterative procedure that estimates a 3D
deformation field between two MR images. At first, large deformations are
estimated using blurred version of the input data. These larger deformations
are then input to subsequent steps where the fit is refined by estimating
smaller deformations on data blurred with a Gaussian kernel with a smaller
FWHM. The final transformation is a set of local translations defined on a bed
of equally spaced nodes that were estimated through the optimization of the
correlation coefficient.

For the purposes of this work we used the regularization parameters optimized 
in Robbins et al. \citep{Robbins2004}.

\todo{link to MNI website here}
\todo{show command line used}

@@

{\bf Dataset evaluated.} 69 1.5T images were randomly selected from the 
{\em ADNI1:Screening 1.5T} standardized dataset.  Demographics for this subset 
are shown in Table X
\todo{FIXME: table reference}.

<<ADNI-69-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl' & RID %in% adni_69_RIDs, 
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics")
@

>>>>

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{figure/MAGeT-figure.png}
  \caption{Diagram of the MAGeT Brain algorithm \label{fig:MAGeT-diagram}}
\end{figure}

@@

\subsubsection{ADNI1 1.5T Screening Dataset}

Clinical, demographic and pre-processed T1-weighted MRI used in this paper were
downloaded by the authors from the ADNI1 database (adni.loni.ucla.edu) between
March 2012 and August 2012. The image dataset download was the "ADNI1:Screening
1.5T" standardized dataset available from ADNI
(http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/)\citep{Wyman2012}.
This image collection contains uniformly preprocessed images which have been
designated to be the "best" after quality control.  All images were acquired
using 1.5T scanners (General Electric Healthcare, Philips Medical Systems or
Siemens Medical Solutions) at multiple sites using the protocol described in
\citep{Jack2008a}.  Representative 1.5T imaging parameters were TR = 2400ms, TI
= 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x 240mm, a
192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel resolution of
1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown in table
\ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

%Representative 3T image parameters were TR = 2300ms, TI = 900ms, TE = 3.0ms,
%flip angle = $8{\circ}$, field of view = 256 x 240mm, a 256 x 256 x 166 matrix
%(x, y, and z directions) yielding a voxel resolution of 1.00 x 1.00 x 1.2
%$mm^3$.

\todo{describe SNT manual segmentations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsubsection{SZ First Episode Patients}

\todo{include image characteristics}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First Episode SZ Demographics Table                                         %
%                                                                             %
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics",
      title="tab:SZFEP_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsubsection{Automatic Normalization and Image Matching and Anatomical
Labeling (ANIMAL)}

The ANIMAL algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the ANIMAL algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in Robbins et al. \citep{Robbins2004},
displayed in table \ref{tab:ANIMAL_params}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                                                           %
%                                                                             %
\begin{table*}[!tbp]
\scriptsize
\caption{ANIMAL Registration Parameters\label{tab:ANIMAL_params}} 
\begin{center}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\end{table*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

{\bf Dataset evaluated.} 69 1.5T images were randomly selected from the 
{\em ADNI1:Screening 1.5T} standardized dataset.  Demographics for this subset 
are shown in Table \ref{tab:ADNI1_xvalidation_demographics}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-69-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics", 
      title="tab:ADNI1_xvalidation_demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
